---
title: "Transcriptional Neoteny in the Human Brain"
author: "Group 7"
format: revealjs
editor: visual
---

# Introduction

Evolution can be explained by developmental timing differences, with neoteny proposed as a mechanism contributing to cognitive traits. 
The study analyzes gene expression during postnatal brain development in humans, chimpanzees, and macaques to explore neotenic shifts.

**Biological Insight**:
- Human prefrontal cortex development shows delayed gene expression, influencing neural plasticity.

**Objective**:
- Analyze datasets using PCA, regression, and heterochrony tests to reveal insights into gene expression shifts.

---

## Materials and Methods

### Data Overview
- **Source**: Somel et al. (2009), GEO datasets (GSE11528, GSE11512, GSE15163).
- **Samples**:
  - Humans: 39 (0–47 years).
  - Chimpanzees: 14 (0–44 years).
  - Macaques: 9 (1–18 years).

**Pipeline**:
1. Data cleaning and augmentation.
2. Principal Component Analysis (PCA).
3. Regression models for age and species effects.
4. Heterochrony tests to detect timing shifts.

---

## Preparing the Data: Cleaning and Augmentation

### Step 1: Data Cleaning
- Addressed missing values and inconsistencies.
- Verified dataset structure and variables.
```{r}
quarto::quarto_render(input = "02_clean.qmd")
```
### Step 2: Augmentation
- Merged datasets across species.
- Derived variables for age and expression scaling.
```{r}
quarto::quarto_render(input = "03_augment_joined.qmd")
```
---

## Decoding Variance: PCA Highlights Developmental Patterns

### Principal Component Analysis
- Identified major sources of expression variance.
### **Key Result**:
- PCA separated species along Components 1 and 2.

**Visualization**:
```{r}
quarto::quarto_render(input = "05_analysis_PCA.qmd")
```
---

## How Much Do Components Explain? Variance Quantified (PVE)

-	Quantified contribution of components to overall variance.

**Visualization**:
```{r}
quarto::quarto_render(input = "12_plot_pve.qmd")
```
---

## PCA Insights: Unique Developmental Trajectories in Humans

### Key Insights:

-	PCA Component 1: Correlates with age-related changes.
-	PCA Component 2: Captures species-specific variation.

Biological Finding:
-	Humans exhibit unique developmental trajectories in PCA space.

**Visualization**:
```{r}
quarto::quarto_render(input = "10_plot_PCA.qmd")
```
---
## From Trends to Insights: Regression Models and Expression Patterns

### Multiple Linear Regression

-	Explored age and species effects on gene expression.

**Visualization**:
```{r}
quarto::quarto_render(input = "06_analysis_multi.qmd")
```
### Cubic Regression

-	Better fit for nonlinear patterns in gene expression.

**Visualization**:
```{r}
quarto::quarto_render(input = "07_analysis_cube.qmd")
```
---

## Comparing Species: Global Trends in Neural Expression

### Overview

-	Compared global expression profiles across species.

### Key Insight:
-	Humans show delayed expression of neural plasticity genes.

**Visualization**:
```{r}
quarto::quarto_render(input = "11_plot_global_expr.qmd")
```
---

## Gene Expression Timing: Detecting Shifts and Neotenic Trends

### Shift Function

-	Analyzed expression timing changes within genes.

**Visualization**:
```{r}
quarto::quarto_render(input = "13_plot_shift_model.qmd")
```
---

## Gene Expression Timing: Detecting Shifts and Neotenic Trends

-	Detected neotenic shifts in 38% of genes.

**Visualization**:
```{r}
quarto::quarto_render(input = "14_plot_heterochrony.qmd")
```
---

## Mapping Gene Expression: Insights Across Evolutionary Scales

### Insights

-	Distribution of genes shows phylogenetic and ontogenetic trends.
-	Neural plasticity genes enriched in neotenic shifts.
---

## Key Findings and Future Directions in Brain Evolution

###	Neotenic Delays:
-	Observed in prefrontal cortex gene expression.
### Evolutionary Significance:
-	Delayed development enhances cognitive capacity.
### Future Directions:
- Functional validation of key genes and pathways.
---
## References

- Somel, M., Liu, X., Tang, L., Yan, Z., Hu, H., Guo, S., … & Khaitovich, P. (2009). Transcriptional neoteny in the human brain. 
Proceedings of the National Academy of Sciences, 106(14), 5743-5748.
---
