---
title: "Augmented Data, organized"
output: html
execute:
  echo: true
---

This script will handle the first part of the augmentation, namely, extracting the data of interest, editing naming and observation format, and joining our two data frames into one.

For this, we will make use of `tidyverse` .

```{r}
#| output: false

library("tidyverse")
```

For the expression data, we need to remove the '\_at' from the gene names. Also, all genes start with 'ENSG' followed by a number, where only the last 6 digits are unique for each gene, and thus we have chosen to remove 'ENS' the non-unique 0s for better readability.

```{r}
#| output: false

# Set path for processed data directory
processed_data_dir <- file.path("..","data")

# Load the expr dataset
expr_data <- read_tsv(file = file.path(processed_data_dir,
                                         "02.1_dat_clean_expr.tsv"))

# Apply above described augmentatitons
expr_data_augment <- expr_data |>
  rename_with(~ sub("00000", "", .), starts_with("ENSG")) |>
  rename_with(~ gsub("ENS|_at", "", .), starts_with("ENSG"))
```

For the phenotype data, we extract the columns 'geo_accession', 'organism_ch1', 'age:ch1', 'sex:ch1.1', 'tissue:ch1.2' an rename the columns - in opposite order.

```{r}
#| output: false

# Load the pheno dataset
pheno_data <- read_tsv(file = file.path(processed_data_dir,
                                         "02.2_dat_clean_pheno.tsv"))

# extract above columns
pheno_data_filter <- pheno_data |>
  rename(sample = geo_accession,
         species = organism_ch1,
         age = `age:ch1`,
         sex = `sex:ch1`,
         tissue = `tissue:ch1`) |>
  select(sample, species, age, sex, tissue)
```

We also change the observations of new 'Age' column to only contain number, substitute Male and Female to M and F, and change the Latin name for the species.

```{r}
#| output: false

pheno_data_augment <- pheno_data_filter |>
  mutate(age = gsub("\\D", "", age) |>
           parse_number(),
         sex = case_when(sex == "Female" ~ "F",
                         sex == "Male" ~ "M"),
         species = case_when(species == "Homo sapiens" ~ "Human",
                             species == "Pan troglodytes" ~ "Chimpanzee",
                             species == "Macaca mulatta" ~ "Macaque"))
```

Finally, we join the two data frames by the shared column 'sample'.

```{r}
#| output: false

pheno_expr_data_joined <- full_join(pheno_data_augment,
                                    expr_data_augment,
                                    by = "sample")
```

We finish by writing a new *03_dat_joined.tsv* file.

```{r}
#| output: false

pheno_expr_data_joined |>
  write_tsv(file = file.path(processed_data_dir,
                             "03_dat_joined.tsv"))
```
