---
title: "Load Data"
output: html
execute:
  echo: true
  output: false
---

This document contains the specifications used for the plots of this project. We use the `RColorBrewer` package to define a set of colors.

```{r}
library("RColorBrewer")
library("ggplot2")
```

Our data contains two main groups of variables, 'species' and 'factors', and one alternating variable that requires distinction by color. We thus use a tricolor palette consisting of the secondary colors purple, green, and orange.

```{r}
tricolor_palette <- c("purple", "green", "orange")
```

Within each secondary color, we define a spectrum to vary within our groups.

```{r}
species_base_colors <- colorRampPalette(brewer.pal(9, "Purples"))(3)

species_colors <- rgb(colorRamp(species_base_colors)
                      (seq(0.2, 0.8,
                           length.out = length(species_base_colors))),
                      maxColorValue = 255)
  
factors_base_colors <- colorRampPalette(brewer.pal(9, "Greens"))(3)

factors_colors <- rgb(colorRamp(factors_base_colors)
                      (seq(0.2, 0.8,
                           length.out = length(factors_base_colors))),
                      maxColorValue = 255)

alternative_base_colors <- colorRampPalette(brewer.pal(9, "Oranges"))(4)

alternative_colors <- rgb(colorRamp(alternative_base_colors)
                      (seq(0.2, 0.8,
                           length.out = length(alternative_base_colors))),
                      maxColorValue = 255)
```

Example plot.

```{r}
combined_colors <- c(species_colors, factors_colors, alternative_colors)
group_labels <- c(
  rep("Species", length(species_colors)),
  rep("Factors", length(factors_colors)),
  rep("Alternative", length(alternative_colors)))

barplot(
  rep(1, length(combined_colors)), 
  col = combined_colors, 
  border = NA, 
  space = 0.2,
  main = "Combined Color Palettes",
  xlab = "",
  ylab = "Color Blocks",
  names.arg = group_labels,
  las = 2,
  cex.names = 0.8)
```
