---
title: "Analysis 1"
output: html
execute:
  echo: true
---

# Load Libraries

In the first analysis we will ...TODO... To do this we will download the `tidyverse` package to use `dplyr`, `tidyr`, and `ggplot2`.

```{r}
#| output: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(broom)
```

First we will reshape the normalized (by age) data to long format for plotting use.

```{r}
normalized_joined_long <- normalized_joined |>
  pivot_longer(cols = starts_with("G"), 
               names_to = "gene", 
               values_to = "expression") |>
  group_by(species, gene) |>
  mutate(min_age = min(normalized_age),
         percent_change = abs((expression - expression[normalized_age == min_age]) / expression[normalized_age == min_age] * 100)) |>
  ungroup()
```

# Plot 1

## Variance Explained in Gene Expression for Sex and Age for each Species

```{r}

```

# Plot 2

## Significant Differences cause by Age or Species Differences in Humans vs both Primates

Assess significant expression change due to normalized age. 0.05 threshold using LM

```{r}
significance <- normalized_joined_long |>
  group_by(gene) |>
  summarize(
    model = list(lm(expression ~ normalized_age)),
    .groups = "drop"
  ) |>
  mutate(
    tidy_model = map(model, tidy)
  ) |>
  unnest(tidy_model) |>
  filter(term == "normalized_age") |>
  select(gene, p_value = p.value) |>
  mutate(
    q_value = p.adjust(p_value, method = "fdr"),
    sig_age = ifelse(q_value < 0.10, 1, 0) # 10% FDR as used in report
  
  )

normalized_joined_long <- normalized_joined_long |>
  left_join(significance |> select(gene, sig_age), by = "gene")
```

Significance tests for expression change due to species using 0.05 threshold in ANOVA.

```{r}
significance <- normalized_joined_long |>
  group_by(gene) |>
  summarize(
    model = list(aov(expression ~ species)),
    .groups = "drop"
  ) |>
  mutate(
    tidy_model = map(model, tidy)
  ) |>
  unnest(tidy_model) |>
  filter(term == "species") |>
  select(gene, p_value = p.value) |>
  mutate(
    q_value = p.adjust(p_value, method = "fdr"),
    sig_sp = ifelse(q_value < 0.10, 1, 0) # 10% FDR as used in report
  )

normalized_joined_long <- normalized_joined_long |>
  left_join(significance |> select(gene, sig_sp), by = "gene")
```

Pie chart showing counts for significance

```{r}
donut_data <- normalized_joined_long |>
  mutate(
    category = case_when(
      sig_age == 1 & sig_sp == 1 ~ "Significant Age & Species",
      sig_age == 0 & sig_sp == 1 ~ "Non-significant Age, Significant Species",
      sig_age == 0 & sig_sp == 0 ~ "Non-significant Age & Species",
      sig_age == 1 & sig_sp == 0 ~ "Significant Age, Non-significant Species"
    )
  ) |>
  count(category) |>
  mutate(percentage = n / sum(n) * 100)
    
ggplot(donut_data, 
       aes(x = 2, 
           y = n, 
           fill = category)) +
  geom_bar(stat = "identity", 
           width = 1, 
           color = "white") +
  coord_polar(theta = "y") +
  theme_void() +
  xlim(0.5, 2.5) +  #donut hole size
  geom_text(aes(
    y = cumsum(n) - n / 2, # position percentages
    label = paste0(round(percentage, 1), "%")
  ), color = "white", 
  size = 4) +
  labs(fill = "Attribute Significance", 
       title = "Significant Change in Gene Expression by Age or Species",
       subtitle = "Significance level = 0.05") +
  theme(plot.title = element_text(hjust = 0.5))
```

# Plot 3

## Percent Change in All Postnatal Expressions

### Data Preparation

FInd the percent change in gene expression relative to the the youngest observation in a given species for each gene.

```{r}
average_change <- normalized_joined_long |>
  group_by(species, sample, normalized_age) |>
  summarise(avg_percent_change = mean(percent_change)) |>
  ungroup()
```

### Plot Creation

The plot will display the percent change in all genes since the youngest observation. The age (x-axis) is normalized, where 1 is the average life expectancy of each species. Values closer to 0 represent the youngest observations.

A sexual maturity line will be added for each of the given ages:

-   Human: 5931.25 days from [Human Development: A Life-Span View](https://books.google.dk/books?id=E-n5E7oyCgoC&pg=PA296&redir_esc=y)

-   Chimpanzee: 2555 days from [WWF](https://wwf.panda.org/discover/knowledge_hub/endangered_species/great_apes/chimpanzees/#:~:text=Although%20chimpanzees%20reach%20sexual%20maturity,until%20they%20are%20over%2060.)

-   Macaque: 1788.5 days from [NIH](https://pmc.ncbi.nlm.nih.gov/articles/PMC10092073/#:~:text=Cynomolgus%20macaque%20sexual%20maturation%20is,for%20CMs%20based%20on%20different)

To calculate where the line should go on the normalized age scale, the values are normalized using the average life expectancy. The normalized sexual maturity ages are as follows:

-   Human: 0.2276

-   Chimpanzee: 0.2125

-   Macaque: 0.3264

"(*D*) The percentage of global expression change relative to newborns. One-hundred percent change was designated as the difference between the youngest and oldest individuals (in humans or chimpanzees) in terms of the summary measure of global expression (see *Materials and Methods*). Each point represents an individual."

```{r}
ggplot(average_change, aes(x = normalized_age, y = avg_percent_change, color = species)) +
  geom_point() +
  labs(title = "Percent Change in all Postnatal Expression by Normalized Age",
       x = "Normalized Age", 
       y = "Average % Change in Gene Expression", 
       color = "Species") +
  theme_minimal() +
  scale_color_manual(values = c("Human" = "maroon2", "Chimpanzee" = "blue2", "Macaque" = "chartreuse4")) +
  geom_vline(xintercept = 0.2276, linetype = "dashed", color = "maroon2", size = 1) +  # Human
  geom_vline(xintercept = 0.2125, linetype = "dashed", color = "blue2", size = 1) +    # Chimpanzee
  geom_vline(xintercept = 0.3264, linetype = "dashed", color = "chartreuse4", size = 1)
```
