---
title: "Heterochrony expression test"
output: html
output-dir: results
execute:
  echo: true
---

```{r}
#| output: false

library("tidyverse")
library("broom")

source("99_proj_func.R")
```

Age normalized, significant expression changes, modeld gene expression trajectories.

```{r}
#| output: false

# Set path for processed data directory
processed_data_dir <- file.path("..","data")

# Load cube analyzed dataset
data_cube <- read_tsv(file = file.path(processed_data_dir,
                                      "06.2_dat_cube.tsv"))
```

```{r}
class_model <- function(data_frame){
  classification <- data_frame |>
    mutate(class = case_when(
      p_value < 0.5 ~ case_when(
        age_shift > 0 ~ "Neogenic",
        age_shift < 0 ~ "Accelerated",
        TRUE ~ "Unclassified"),
      TRUE ~ "Unclassified"))
  
  neogenic <- classification |>
    filter(class == "Neogenic") |>
    pull(gene)
  
  accelerated <- classification |>
    filter(class == "Accelerated") |>
    pull(gene)

  unclassified <- classification |>
    filter(class == "Unclassified") |>
    pull(gene)
  
  # Return
  list(
    neogenic = neogenic,
    accelerated = accelerated,
    unclassified = unclassified)}
```

```{r}
test1 <- test |> class_model()

```

```{r}
human_chimp_shift <- human_chimp_filtered |>
  group_by(gene) |>
  nest() |>
  mutate(
    shift_modeling = map(.x = data,
                         .f = ~ shift_model(data_frame = .x))) |>
  mutate(
    age_shift = map_dbl(shift_modeling, "age_shift"),
    expr_shift = map_dbl(shift_modeling, "expr_shift"),
    p_value = map_dbl(shift_modeling, "p_value"),
    rss_diff = map_dbl(shift_modeling, "rss_diff")) |>
  dplyr::select(-shift_modeling)
```

```{r}
human_chimp_class <- human_chimp_shift |> 
  mutate(class = case_when(
    p_value < 0.5 ~ case_when(
      age_shift > 0 ~ "Neogenic",
      age_shift < 0 ~ "Accelerated",
      TRUE ~ "Unclassified"),
    TRUE ~ "Unclassified"))
```

**Model Gene Expression Trajectories**:

-   Fit expression trajectories across ages for humans and chimpanzees, assuming similar ontogenetic shapes.

-   Use models (e.g., linear regression) to describe age-related expression changes for each gene.

**Estimate Timing Differences**:

-   Introduce an "age-shift" or "expression-shift" between species to align the expression curves.

-   Evaluate whether these shifts significantly improve the model fit (e.g., F-test or likelihood ratio test).

**Classify Heterochronic Genes**:

-   Based on the direction and magnitude of shifts:

    -   **Human Neoteny**: Human expression resembles younger chimpanzees.

    -   **Human Acceleration**: Human expression resembles older chimpanzees.

    -   **Chimpanzee Neoteny/Acceleration**: Reciprocal classification for chimpanzees.

```{r}
# Extract delayed and accelerated genes for Human vs. Chimpanzee
human_chimp_neogenic <- human_chimp_class |> 
  filter(class == "Neogenic") |> 
  pull(gene)

human_chimp_accelerated <- human_chimp_class |> 
  filter(class == "Accelerated") |> 
  pull(gene)

```
