---
title: "Cube Regression Analysis"
output: html
execute:
  echo: true
  output: false
---

#### Load

For this analysis, we follow the description found in the appendix of *Somel et al*. It describes that global gene expression is calculated using the `MASS` package.

```{r}
library("tidyverse")
library("broom")
library("MASS")
```

Load the data set.

```{r}
# Set path for processed data directory
processed_data_dir <- file.path("..",
                                "data")

# Load the dataset
data <- read_tsv(file = file.path(processed_data_dir,
                                         "03_dat_joined.tsv"),
                 show_col_types = FALSE)
```

#### Cube regression

Fit polynomial (cubic) regression model to each gene and species, calculate `r_squared` and `p_value` using ANOVA test. As described by appendix, age is usually log transformed for this type of analysis.

```{r}
# Create long data frame
data_cube <- data |>
  pivot_longer(
    cols = starts_with("G"),
    names_to = "gene",
    values_to = "expr") |>
  mutate(log_expr = log2(expr + 1)) |>
  group_by(gene,
           species) |>
  nest()

# Perform cubic regression and ANOVA analysis
data_cube_reg <- data_cube |>
  mutate(
    cubic_model = map(.x = data,
                      .f = ~ lm(log_expr ~ poly(log_age, 3),
                                data = .)),
    r_squared = map_dbl(.x = cubic_model,
                        .f = ~ summary(.) |>
                          pluck("r.squared")),
    p_value = map_dbl(.x = cubic_model,
                      .f = ~ anova(.) |>
                        tidy() |>
                        filter(term == "poly(log_age, 3)") |> 
                        pull(p.value)))
```

Extract coefficients from the cube regression models.

```{r}
data_cube_coef <- data_cube_reg |>
  mutate(coefficients = map(.x = cubic_model,
                            .f = ~ tidy(.x))) |>
  mutate(intercept = map_dbl(.x =coefficients,
                             .f = ~ filter(.,
                                           term == "(Intercept)") |>
                               pull(estimate)),
         poly1 = map_dbl(.x = coefficients,
                         .f = ~ filter(.,
                                       term == "poly(log_age, 3)1") |>
                           pull(estimate)),
         poly2 = map_dbl(.x = coefficients,
                         .f = ~filter(.,
                                      term == "poly(log_age, 3)2") |>
                           pull(estimate)),
         poly3 = map_dbl(.x = coefficients,
                         .f = ~filter(.,
                                      term == "poly(log_age, 3)3") |>
                           pull(estimate))) |>
  unnest(cols = c(data)) |>
  dplyr::select(-cubic_model,
                -coefficients,
                -sample,
                -sex,
                -log_expr) |>
  ungroup()
```

#### Significant Genes

Identify and filter genes displaying a significant change in expression level.

```{r}
# Identify significant genes
significant_genes <- data_cube_reg |> 
  filter(p_value < 0.05 & r_squared > 0.2) |>
  pull(gene) |> 
  unique()

# Extract genes
significant_data <- data_cube |>
  filter(gene %in% significant_genes) |> 
  unnest(cols = c(data)) |>
  dplyr::select(-log_expr) |>
  ungroup() |>
  pivot_wider(names_from = gene,
              values_from = expr)

```

To obtain a one-dimensional summary measure of expression variation, the matrix of Euclidean distances among the samples is calculated.

```{r}
#| warning: false

dist_matrix <- significant_data |>
  dplyr::select(-species,
                -age,
                -log_age,
                -sample) |>
  dist() |>
  as.matrix()
```

Apply monotonic regression to collapse the distance into a single dimension.

```{r}
# Apply monotonic regression
mds_result <- isoMDS(dist_matrix,
                     k = 1)

# Add result to data
significant_data <- significant_data |> 
  mutate(global_expr = mds_result |>
           pluck("points") |>
           as.numeric())
```

#### Global Expression

Calculate the percent change in global expression, with 0% and 100% being equal to the expression levels for the youngest and the oldest sample, respectively.

```{r}
scaled_data <- significant_data |> 
  group_by(species) |> 
  mutate(
    min_global = filter(pick(everything()),
                        age == min(age)) |>
      pull(global_expr),
    max_global = filter(pick(everything()),
                        age == max(age)) |>
      pull(global_expr),
    percent_global =
      (global_expr - min_global) / c(max_global - min_global) * 100) |>
  ungroup()
```

#### Save

Write results to .*tsv* files.

```{r}
scaled_data |> write_tsv(file = file.path(processed_data_dir,
                                          "07.1_dat_global_expr.tsv"))

data_cube_coef |> write_tsv(file = file.path(processed_data_dir,
                                             "07.2_dat_cube.tsv"))
```

