---
title: "Heterochrony expression test"
output: html
execute:
  echo: true
  output: false
---

#### Load

For the heterochrony test, we use the functions described in *99_proj_func.R*.

Load libraries.

```{r}
library("tidyverse")
library("broom")

source("15_proj_func.R")
```

Load the data sets.

```{r}
#| output: false

# Set path for processed data directory
processed_data_dir <- file.path("..","data")

# Load cube analyzed dataset
data_cube <- read_tsv(file = file.path(processed_data_dir,
                                      "07.2_dat_cube.tsv"),
                      show_col_types = FALSE)
```

#### **Estimate Timing Differences**

-   Introduce an "age-shift" or "expression-shift" between species to align the expression curves.

-   Evaluate whether these shifts significantly improve the model fit using F-test.

```{r}
#| warning: false

# Human vs chimp and reciprocal
human_chimp_shift <- data_cube |>
  apply_shift_model(original_species = "Human",
                    compared_species = "Chimpanzee")

chimp_human_shift <- data_cube |>
  apply_shift_model(original_species = "Chimpanzee",
                    compared_species = "Human")

# Human vs macaque and reciprocal
human_mac_shift <- data_cube |>
  apply_shift_model(original_species = "Human",
                    compared_species = "Macaque")

mac_human_shift <- data_cube |>
  apply_shift_model(original_species = "Macaque",
                    compared_species = "Human")
```

#### **Classify Heterochronic Genes**

Based on the direction and magnitude of shifts:

-   **Neotenic**: Original species' expression resembles younger target species. Defined as a positive shift in age or expression.

-   **Accelerated**: Original species' expression resembles older target species. Defined as a negative shift in age or expression.

-   **Undefined**: Unable to determine from model. Handles `NAs`.

Genes that do not display a significant improvement in fit by the shift model, are classified as **Non-significant**.

```{r}
# Human vs chimp and reciprocal
human_chimp_class <- human_chimp_shift |>
  class_model()

chimp_human_class <- chimp_human_shift |>
  class_model()

# Human vs macaque and reciprocal
human_mac_class <- human_mac_shift |>
  class_model()

mac_human_class <- mac_human_shift |>
  class_model()
```

Add classifications to shift data frame.

```{r}
# Human vs chimp and reciprocal
human_chimp_table <- human_chimp_class |>
  enframe(name = "classification",
          value = "genes") |>
  unnest(cols = c(genes)) |>
  left_join(human_chimp_shift,
            by = c("genes" = "gene"))

chimp_human_table <- chimp_human_class |>
  enframe(name = "classification",
          value = "genes") |>
  unnest(cols = c(genes)) |>
  left_join(chimp_human_shift,
            by = c("genes" = "gene"))

# Human vs macaque and reciprocal
human_mac_table <- human_mac_class |>
  enframe(name = "classification",
          value = "genes") |>
  unnest(cols = c(genes)) |>
  left_join(human_mac_shift,
            by = c("genes" = "gene"))

mac_human_table <- mac_human_class |>
  enframe(name = "classification",
          value = "genes") |>
  unnest(cols = c(genes)) |>
  left_join(mac_human_shift,
            by = c("genes" = "gene"))
```

Calculate proportion of genes for each classification and combine data into two data frames containing species interaction and their reciprocal.

```{r}
human_chimp_recip <- human_chimp_table  |>
  count(classification) |>
  mutate(hum_proportion = n / sum(n) * 100) |>
  rename(hum_n = n) |>
  left_join(chimp_human_table |>
              count(classification) |>
              mutate(chimp_proportion = n / sum(n) * 100) |>
              rename(chimp_n = n),
            by = join_by(classification))

human_mac_recip <- human_mac_table  |>
  count(classification) |>
  mutate(hum_proportion = n / sum(n) * 100) |>
  rename(hum_n = n) |>
  left_join(mac_human_table |>
              count(classification) |>
              mutate(mac_proportion = n / sum(n) * 100) |>
              rename(mac_n = n),
            by = join_by(classification))
```

#### Save

Write results to .*tsv* files.

```{r}
human_chimp_recip |>
  write_tsv(file = file.path(processed_data_dir,
                             "08.1_hum_chi_class.tsv"))
human_mac_recip |>
  write_tsv(file = file.path(processed_data_dir,
                             "08.2_hum_mac_class.tsv"))

human_chimp_shift |>
  dplyr::select(-shift_p_value,
                -rss_diff) |>
  unnest(cols = c(data)) |>
  write_tsv(file = file.path(processed_data_dir,
                             "08.3_hum_chi_shift.tsv"))
human_mac_shift |>
  dplyr::select(-shift_p_value,
                -rss_diff) |>  
  unnest(cols = c(data)) |>
  write_tsv(file = file.path(processed_data_dir,
                             "08.4_hum_mac_shift.tsv"))
```
