---
title: "Principle Component Analysis, plots"
output: html
output-dir: results
execute:
  echo: true
  output: false
---

For the principal component analysis (PCA) we want to visualize the data projected into the principal components, along the proportion of variance explained (PVE) for each component.

```{r}
library("tidyverse")
library("broom")
library("patchwork")

source("99_proj_func.R")
```

We load the data sets.

```{r}
# Set path for processed data directory
processed_data_dir <- file.path("..","data")

# Load the data sets
proj_data <- read_tsv(file = file.path(processed_data_dir,
                                         "05.2_dat_pca_proj.tsv"),
                      show_col_types = FALSE)
pve_data <- read_tsv(file = file.path(processed_data_dir,
                                         "05.3_dat_pve_all.tsv"),
                     show_col_types = FALSE)
color_palette <- read_tsv(file = file.path(processed_data_dir,
                                         "04_col_pal.tsv"),
                     show_col_types = FALSE)
```

```{r}
age_color <- color_palette |>
  filter(group == "Attributes") |>
  slice(1) |>
  pull(color)
sex_color <- color_palette |>
  filter(group == "Attributes") |>
  slice(2) |>
  pull(color)
spec_color <- color_palette |>
  filter(group == "Attributes") |>
  slice(3) |>
  pull(color)
```

```{r}
pve_data_prep <- pve_data |>
  group_by(PC) |>
  nest() |>
  mutate(sum_var_expl_pc = map_dbl(.x = data,
                                   .f = ~ .x |>
                                     pull(var_expl_pc)|>
                                     sum())) |>
  unnest(cols = c(data)) |>
  ungroup() |>
  mutate(plt_var_expl = (var_expl_pc / sum_var_expl_pc) * variance_explained) |>
  mutate(attribute = case_when(attribute == "log_age" ~ "Age",
                            attribute == "species_num" ~ "Species",
                            attribute == "sex_num" ~ "Sex")) |>
  mutate(PC = factor(PC,
                     levels = unique(PC)))

```

First we plot the variance explained for each principal component, colored by the PVE by attributes.

```{r}
plot_pc_pve <- pve_data_prep |>
  ggplot(aes(x = PC,
             y = plt_var_expl,
             fill = attribute)) + 
  geom_bar(stat = "identity",
           position = "stack") +
  geom_hline(yintercept = 1,
             linetype = "dashed") +
  labs(
    x = "Principal Component",
    y = "Variance Explained (%)",
    fill = "Attribute",
    title = "Variance Explained by Principal Components and Attributes") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.text.x = element_blank()) +
  scale_fill_manual(values = color_palette |>
                      filter(group == "Attributes") |>
                      pull(color))

plot_pc_pve
```

The projection data is visualized by plotting the data onto the first two principal components, along with details about age, species, and sex.

```{r}
plot_1_3_proj <- proj_data |>
  pca_proj_plot(principal_component_1 = "PC1",
                principal_component_2 = "PC3")

plot_2_3_proj <- proj_data |>
  pca_proj_plot(principal_component_1 = "PC2",
               principal_component_2 = "PC3")

plot_6_9_proj <- proj_data |>
  pca_proj_plot(principal_component_1 = "PC6",
                principal_component_2 = "PC9")


plot_1_3_proj

plot_2_3_proj

plot_6_9_proj
```

For the variance calculation, we make a plot of the cumulative along side the variance explained over all principal components.

We plot the variance explained for each principal component.

```{r}
proj_plots <- plot_1_3_proj +
  plot_2_3_proj +
  plot_6_9_proj

combined_plot <- plot_pc_pve +
  inset_element(proj_plots, 0.1, 0.7, 1, 1) +
  plot_annotation(
    title = "PCA reveals clear trends of dependencies of expression on attributes",
    subtitle = "The main PVE is by the attributes 
    <span style='color:age_color;'>age</span>
    and
    <span style='color:spec_color;'>species</span>...",
    caption = "...whereas <span style='color:sex_color;'>sex</span> is mainly represented in the lower ranking principal components.",
    theme = theme(
      plot.title = element_text(size = 16,
                                face = "bold",
                                hjust = -1),
      plot.subtitle = element_text(size = 10,
                                   hjust = -1),
      plot.caption = element_text(size = 10,
                                  hjust = 1)))


combined_plot
```

Write final plot to *.png* file.

```{r}
ggsave(file = file.path(processed_data_dir,
                        "08_test.png"),
       plot = combined_plot,
       width = 6,
       height = 4,
       dpi = 300)
```
