---
title: "Heterochrony, plots"
format: html
output-dir: results
execute:
  echo: true
  output: false
---

We begin with loading the necessary libraries for plot formation:

```{r}
library(tidyverse)
```

Thereafter, we load the data set from the heterochrony analysis and for our chosen graphic colours:

```{r}
# Set path for processed data directory
# blank

# Load the data sets
# blank

color_palette <- read_tsv(file = file.path(processed_data_dir,
                                         "04_col_pal.tsv"),
                     show_col_types = FALSE)
```

```{r}
# Test dataset
test_set <- data.frame(
  Gene = c("GeneA", "GeneB", "GeneC", "GeneD", "GeneE", "GeneF", "GeneG", "GeneH","GeneQ", "Gene1", "Gene2", "Gene3", "Gene4", "Gene5", "Gene6","Gene7"),
  Species = c("Human", "Human", "Monkey", "Monkey", "Monkey", "Monkey","Human","Human","Monkey","Monkey","Human","Monkey","Chimp","Chimp","Chimp","Chimp"),
  Association = c("Neotenic", "Neotenic", "Neotenic", "Accelerated", "Neotenic", "Accelerated", "Accelerated", "Accelerated", "Accelerated","Neotenic","Neotenic", "Accelerated", "Neotenic", "Accelerated", "Neotenic","Accelerated"),
  Tissue = c("Forebrain", "Forebrain", "Forebrain", "Midbrain", "Forebrain", "Midbrain", "Forebrain", "Midbrain", "Midbrain", "Midbrain", "Midbrain", "Forebrain","Midbrain","Forebrain","Forebrain","Midbrain")
)

# Display the dataset
print(test_set)
```

We wish to visualize the percentages of neotenic and accelerated genes from the prefrontal cortex for humans, chimps, and macaques, as well as for the brain tissue caudate nucleus for humans only. Therefore we start by argumentation of the heterochrony data from our earlier analysis, to find the percentages:

```{r}
# Condence to a new dataset with % catagory
percent_phyloontogenic <- test_set |>
  group_by(Species, Association, Tissue) |>
  summarise(Count = n()) |>
  mutate(Percentage = (Count / sum(Count))*100)

print(percent_phyloontogenic)
```

We now can visualize the percentages of neotenic and accelerated genes:

```{r}

#| label: Phyloontogentic_plot1


ggplot(data = percent_phyloontogenic,
       mapping = aes(y= Species,
                     x= Percentage,
                     fill = Association)) +
  geom_bar(stat = "identity",
           position=position_dodge(),
           color = "black") +
    facet_wrap(~Tissue) +
  theme_minimal() +
  theme(legend.position="none",
        plot.title.position = "plot",
        plot.subtitle = element_textbox_simple(),
        text=element_text(size=12,  family="momo"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  labs(title = "Gene Distribution", 
       subtitle = "by Phyloontogentic Catagories: <span style='color:#DE580C'>Accelerated</span> vs. <span style='color:#9E3706'>Neotenic</span>",
       y = NULL,
       x = "Genes Assigned to Categories (%)")+
  scale_fill_manual(values = color_palette |>
                      filter(group == "Alternative") |>
                      slice_tail(n=2) |>
                      pull(color))
```

Finally, we now write our final plot to a .png file:

```{r}
# ggsave(file = FILL IN,
#        plot = combined,
#        width = 6,
#        height = 4,
#        dpi = 300)
```

TO BE REMOVED

.................................................

The proportion of gray-matter-specific genes among human-neotenic genes and 'other', by tissue type (caudate nucleus, prefrontal cortex)

```{r}
# Condence to a new dataset with % grey matter genes caragory
percent_grey <- test_set |>
  filter(Species == "Human") |>
  group_by(Association, Tissue) |>
  summarise(Count = n()) |>
  mutate(Percentage = (Count / sum(Count))*100)

print(percent_grey)
```

```{r}
ggplot(data = percent_grey,
       mapping = aes(y= Tissue,
                     x= Percentage,
                     fill = Association)) +
  geom_bar(stat = "identity",
           position=position_dodge(),
           color = "black") +
  scale_fill_manual(values = c("red","blue"),
                    labels = c("Accelerated","Neotenic"))+
  theme_minimal() +
  theme(legend.position="none",
        plot.title.position = "plot",
        plot.subtitle = element_textbox_simple(),
        text=element_text(size=12,  family="momo"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  labs(title = "Proportion of Gray-Matter-Specific Genes", 
       subtitle = "by Phyloontogentic Catagories: <span style='color:#ff0000'>Accelerated</span> vs. <span style='color:#0000FF'>Neotenic</span>",
       y = NULL,
       x = "%")
```
