---
title: "MLC Runs"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(dplyr)
library(broom)
```

For every gene, multivariable regression is done to determine whether any of the two factors, age, and species, are significant to change in the expression of that particular gene. Later, the proportion of significant genes will be presented.

Since one of the factors (species) is categorical, the first step is to represent these variables with numeric values, 0 and 1. The code below also extracts the age of all samples.

```{r}
processed_data_dir <- file.path("..","data")
pheno_expr_data_joined <- read_tsv(file = file.path(processed_data_dir,
                                         "03_dat_joined.tsv"))
```

```{r}
mlc_data <- pheno_expr_data_joined |> 
  mutate(human_chimp_species = case_when(
    grepl("Human", species) ~ 0,
    grepl("Chimpanzee", species) ~ 1,
    TRUE ~ NA_real_
  ))
mlc_data <- mlc_data |>
  mutate(human_maca_species = case_when(
    grepl("Human", species) ~ 0,
    grepl("Macaque", species) ~ 1,
    TRUE ~ NA_real_
  ))
#mlc_data
```

Process the data by performing the necesary statistical tests. The code returns a dataframe with information on which genes are predicted with great significance by age or species.

```{r}
aug_mlc_stats <- mlc_data |>
  dplyr::select(-sample) |>
  dplyr::select(-species) |>
  dplyr::select(-sex) |>
  map_dfr(~ {
    
    model_chimp <- lm(. ~ age + human_chimp_species, data = mlc_data)
    model_maca <- lm(. ~ age + human_maca_species, data = mlc_data)
    
    # Extract the F-statistic p-values
    f_stat_pvalue_chimp <- model_chimp |>
      glance() |>
      pull(p.value)
    f_stat_pvalue_maca <- model_maca |>
      glance() |> 
      pull(p.value)
    
    # Extract the t-statistic p-values
    t_stat_pvalues_chimp <- model_chimp |>
      tidy() |>
      filter(term %in% c("age", "human_chimp_species")) |>
      dplyr::select(term, p.value) |>
      pivot_wider(names_from = term, 
                  values_from = p.value,
                  names_prefix = "t_stat_pvalue_chimp_")
    
    t_stat_pvalues_maca <- model_maca |>
      tidy() |>
      filter(term %in% c("age", "human_maca_species")) |>
      dplyr::select(term, p.value) |>
      pivot_wider(names_from = term,
                  values_from = p.value,
                  names_prefix = "t_stat_pvalue_maca_")
    
    # Determine if age and species are significant predictors
    age_significant_chimp <- if_else(t_stat_pvalues_chimp$t_stat_pvalue_chimp_age < 0.05, "yes", "no")
    species_significant_chimp <- if_else(t_stat_pvalues_chimp$t_stat_pvalue_chimp_human_chimp_species < 0.05, "yes", "no")
    
    age_significant_maca <- if_else(t_stat_pvalues_maca$t_stat_pvalue_maca_age < 0.05, "yes", "no")
    species_significant_maca <- if_else(t_stat_pvalues_maca$t_stat_pvalue_maca_human_maca_species < 0.05, "yes", "no")
    
    # Combine the results into a single dataframe
    tibble(
      f_stat_pvalue_chimp = f_stat_pvalue_chimp,
      f_stat_pvalue_maca = f_stat_pvalue_maca,
      t_stat_pvalues_chimp,
      t_stat_pvalues_maca,
      age_significant_chimp = age_significant_chimp,
      species_significant_chimp = species_significant_chimp,
      age_significant_maca = age_significant_maca,
      species_significant_maca = species_significant_maca
    )
  })

#aug_mlc_stats
```

Find the fraction of all genes that are significantly predicted by age, species or both.

```{r}
aug_prop_significant <- aug_mlc_stats |>
  summarise(across(ends_with("_significant_chimp"),
                   ~ sum(. == "yes", na.rm = TRUE) / n() * 100,
                   .names = "prop_{.col}"),
            across(ends_with("_significant_maca"),
                   ~ sum(. == "yes", na.rm = TRUE) / n() * 100,
                   .names = "prop_{.col}"),
            prop_both_significant_chimp = sum(age_significant_chimp == "yes" & species_significant_chimp == "yes", na.rm = TRUE) / n() * 100,
            prop_both_significant_maca = sum(age_significant_maca == "yes" & species_significant_maca == "yes", na.rm = TRUE) / n() * 100,
            prop_only_age_significant_chimp = sum(age_significant_chimp == "yes" & species_significant_chimp == "no", na.rm = TRUE) / n() * 100,
            prop_only_species_significant_chimp = sum(age_significant_chimp == "no" & species_significant_chimp == "yes", na.rm = TRUE) / n() * 100,
            prop_only_age_significant_maca = sum(age_significant_maca == "yes" & species_significant_maca == "no", na.rm = TRUE) / n() * 100,
            prop_only_species_significant_maca = sum(age_significant_maca == "no" & species_significant_maca == "yes", na.rm = TRUE) / n() * 100,
            prop_none_significant_chimp = sum(age_significant_chimp == "no" & species_significant_chimp == "no", na.rm = TRUE) / n() * 100,
            prop_none_significant_maca = sum(age_significant_maca == "no" & species_significant_maca == "no", na.rm = TRUE) / n() * 100)

aug_prop_significant

#aug_prop_significant
```

Create a new data frame that contains all proportions of interest, that can be used by the script downstream.

```{r}
mlc_results <- tibble(category = c("Age+/Sp+", "Age+/Sp-", "Age-/Sp+", "Age-/Sp-"),
                      human_chimp = c(pull(aug_prop_significant, var = "prop_both_significant_chimp"),
                                      pull(aug_prop_significant, var="prop_only_age_significant_chimp"),
                                      pull(aug_prop_significant, var = "prop_only_species_significant_chimp"),
                                       pull(aug_prop_significant, var = 
"prop_none_significant_chimp")),
                      human_macaque = c(pull(aug_prop_significant, var = "prop_both_significant_maca"),
                                      pull(aug_prop_significant, var="prop_only_age_significant_maca"),
                                      pull(aug_prop_significant, var = "prop_only_species_significant_maca"),
                                       pull(aug_prop_significant, var = 
"prop_none_significant_maca")))


mlc_results
```

Results are documented, and will be used later for visualizing.

```{r}
aug_prop_significant |>
  write_tsv(file = file.path(processed_data_dir,
                             "06.1_dat_mlc_results.tsv"))
mlc_results |> 
  write_tsv(file= file.path(processed_data_dir,
                            "06.2_dat_mlc_proportions.tsv"))
```
